Learning semi-Markovian DAGs with flow-based VAE

### 2023-12-11 Tex/slide.tex
slide 최종

\begin{frame}{DAG and SEM}
    \begin{itemize}
        \item Directed Acyclic Graph(DAG) $\mathcal G = (V, E)$ with $V = \{1, \dots, m\}$
        \item Linear Structural Equation Model(SEM) is defined as
        \begin{equation*}
            X_i = \sum_{j \in \text{Pa}(i)} \beta_{ij} X_j + \epsilon_i
        \end{equation*}
        where $\epsilon_i$ is a noise variable.
        \item Matrix form can be written as
        \begin{equation}
            \mathbf X = \mathbf A^T \mathbf X + \mathbf \epsilon
        \end{equation}
        where $\mathbf A$ is an adjacency matrix of $\mathcal G$ and $\mathbf \epsilon$ is a noise vector.
    \end{itemize}
\end{frame}

\begin{frame}{DAG and SEM(Cont'd)}
    \begin{itemize}
        \item If $I-\mathbf A^T$ is nonsingular, SEM can be written as
        \begin{equation}
            \mathbf X = (\mathbf I - \mathbf A^T)^{-1} \mathbf \epsilon
        \end{equation}
        \item Usually we assume $\mathbf \epsilon \sim \mathcal N(\mathbf 0, \mathbf \Sigma)$ and
        \begin{equation}
            \mathbf X \sim \mathcal N(\mathbf 0, (\mathbf I - \mathbf A^T)^{-1} \mathbf \Sigma (\mathbf I - \mathbf A)^{-1})
        \end{equation}
    \end{itemize}
\end{frame}

\begin{frame}{Semi-Markovian DAG}
    \begin{itemize}
        \item Semi-Markovian DAG is a DAG where the noise variables have dependencies \footcite{Shipster2006semimarkovian}
        \item The noise vector $\mathbf \epsilon$ is a semi-Markovian noise vector if
        \begin{equation}
            \Sigma \in \mathcal{W}(G) := \{ \Sigma \in \mathbb R^{m\times m} : \Sigma_{ij} = \Sigma_{ji} = 0 \text{ if } i \notin \text{sib}(j) \}
        \end{equation}
        where $\text{sib}(j) = \{i : i \sim j \text{ in } G\}$ \footcite{wang2017empirical}
    \end{itemize}
\end{frame}

\begin{frame}{DAG Learning}
    \begin{itemize}
        \item DAG learning is a problem of estimating $\mathcal G$ from $\mathbf X$
        \item NP-hard problem \footcite{ChickeringNPhard}
        \item Score-based methods and constraint-based methods are used
        \item NOTEARS : conversion of DAG learning to continuous optimization problem \footcite{zheng2018dags}
        \begin{equation}
            tr(A)
        \end{equation}
    \end{itemize}
\end{frame}

@@

\begin{frame}{Linear IAF}
    \begin{itemize}
        \item Inverse Autoregressive Flow(IAF) is a normalizing flow that uses autoregressive transformation \footcite{kingma2016iaf}
        \item Linear IAF is a special case of IAF where $f_t$ is a linear transformation
        \begin{equation}
            \bf z_t = \bf \mu_t + \bf \sigma_t \odot \bf z_{t-1}.
        \end{equation}
        \item The probability density function of $z_T$ is
        \begin{equation}
            \mathbf{z}_T = \mathbf{L(x)} \cdot \mathbf{z}_0.
        \end{equation}
        where $\mathbf{L(x)}$ is a lower triangular matrix from encoder network.
        \item We can use more flexible prior distribution by using normalizing flow.
    \end{itemize}
\end{frame}

@@

\begin{frame}{Learning}
    \begin{itemize}
        \item We use the same optimization procedure as DAG-GNN \footcite{yu2019daggnn}.
        \item The optimization procedure is minimizing the following loss function:
        \begin{equation}
            \mathcal{L}(A, W, \lambda) =-\mathcal{L}_{\mathrm{ELBO}} + \tau \Vert A \Vert_1 + \lambda h(A) + \frac{c}{2} |h(A)|^2.
        \end{equation}
        \item The second term is the L1 regularization term, which encourages sparsity of the adjacency matrix.
        \item The third and fourth terms are the augmented Lagrangian terms.
        \item Gradually increasing the value of $c$ during the training process results the acyclicity constraint to zero.
    \end{itemize}
\end{frame}

@@

>>>>

\begin{frame}{DAG and SEM}
    \begin{itemize}
        \item Directed Acyclic Graph(DAG) $\mathcal G = (V, E)$ with $V = \{1, \dots, m\}$ is used to represent a causal structure \footcite{pearl_2009}
        \item Linear Structural Equation Model(SEM) is defined as
        \begin{equation*}
            X_i = \sum_{j \in \text{Pa}(i)} \beta_{ij} X_j + \epsilon_i
        \end{equation*}
        where $\epsilon_i$ is a noise variable.
        \item Matrix form can be written as
        \begin{equation}
            \mathbf X = \mathbf A^T \mathbf X + \mathbf \epsilon
        \end{equation}
        where $\mathbf A$ is an adjacency matrix of $\mathcal G$ and $\mathbf \epsilon$ is a noise vector.
    \end{itemize}
\end{frame}

\begin{frame}{DAG and SEM(Cont'd)}
    \begin{itemize}
        \item If $(\mathbf I-\mathbf A^T)$ is nonsingular, SEM can be written as
        \begin{equation}
            \mathbf X = (\mathbf I - \mathbf A^T)^{-1} \mathbf \epsilon
        \end{equation}
        \item Usually we assume $\mathbf \epsilon \sim \mathcal N(\mathbf 0, \mathbf \Sigma)$ and
        \begin{equation*}
            \mathbf X \sim \mathcal N(\mathbf 0, (\mathbf I - \mathbf A^T)^{-1} \mathbf \Sigma (\mathbf I - \mathbf A)^{-1})
        \end{equation*}
    \end{itemize}
\end{frame}

\begin{frame}{Semi-Markovian DAG}
    \begin{itemize}
        \item Semi-Markovian DAG is a DAG where the noise variables have dependencies \footcite{Shipster2006semimarkovian}
        \item The noise vector $\mathbf \epsilon$ is a semi-Markovian noise vector if
        \begin{equation}
            \Sigma \in \mathcal{W}(G) := \{ \Sigma \in \mathbb R^{m\times m} : \Sigma_{ij} = \Sigma_{ji} = 0 \text{ if } i \notin \text{sib}(j) \}
        \end{equation}
        where $\text{sib}(j) = \{i : i \sim j \text{ in } G\}$ \footcite{wang2017empirical}
        \item Learn cyclic adjacency matrix or acyclic adjacency matrix
    \end{itemize}
\end{frame}

\begin{frame}{DAG Learning}
    \begin{itemize}
        \item DAG learning is a problem of estimating $\mathcal G$ from $\mathbf X$
        \item NP-hard problem \footcite{ChickeringNPhard}
        \item Score-based methods \footcite{Spirtes2000} and constraint-based methods \footcite{ChickeringGES} are used
        \item NOTEARS\footcite{zheng2018dags}: Combinatorial optimization to Continuous optimization problem
        \begin{equation*}
            \min_{\mathbf{A}} F(A) \text{ subject to } G(A) \in \mathbb{D} \Rightarrow 
            \min_{\mathbf{A}} F(A) \text{ subject to } h(A)=0
        \end{equation*}
        where $\mathbb D$ denotes the discrete space of DAGs on $m$ nodes and $h(A)=0$ iff $G(A)$ is acyclic.
    \end{itemize}
\end{frame}

\begin{frame}{DAG Learning(Cont'd)}
    \begin{block}{Theorem}
        A matrix $\mathbf A$ is the adjacency matrix of a DAG if and only if
        \begin{equation*}
            h(A) = \rm{tr}(\exp(\mathbf A \circ \mathbf A)) = m
        \end{equation*}
        where $\circ$ denotes the Hadamard product.
    \end{block}
    \begin{block}{Theorem}
        A matrix $\mathbf A$ is the adjacency matrix of a DAG if and only if
        \begin{equation*}
            \rm{tr}[(I + \alpha \mathbf A \circ \mathbf A)^m] = m
        \end{equation*}
        for any $\alpha>0$.
    \end{block}
\end{frame}

@@

\begin{frame}{Linear IAF}
    \begin{itemize}
        \item Inverse Autoregressive Flow(IAF) is a normalizing flow that uses autoregressive transformation \footcite{kingma2016iaf}
        \item Linear IAF is a special case of IAF where $f_t$ is a linear transformation
        \begin{equation}
            \bf z_t = \bf \mu_t + \bf \sigma_t \odot \bf z_{t-1}.
        \end{equation}
        \item Transformation of the linear IAF can be written as
        \begin{equation}
            \mathbf{z}_T = \mathbf{L(x)} \cdot \mathbf{z}_0.
        \end{equation}
        where $\mathbf{L(x)}$ is a lower triangular matrix from encoder network.
        \item We can use more flexible prior distribution by using normalizing flow.
        \item When $\mathbf z_0$ has diagonal covariance, $\mathbf z_T$ then has full covariance matrix.
    \end{itemize}
\end{frame}

@@

\begin{frame}{Learning}
    \begin{itemize}
        \item The optimization procedure is minimizing the following loss function:
        \begin{equation}
            \mathcal{L}(A, W, \lambda) =-\mathcal{L}_{\mathrm{ELBO}} + \tau \Vert A \Vert_1 + \lambda h(A) + \frac{c}{2} |h(A)|^2.
        \end{equation}
        \item The second term is the L1 regularization term, which encourages sparsity of the adjacency matrix.
        \item $\tau=0.1 \times ({m\over 50})^2$ works well in our experiment setting.
        \item The third and fourth terms are the augmented Lagrangian terms.
        \item Gradually increasing the value of $c$ during the training process results the acyclicity constraint to zero.
    \end{itemize}
\end{frame}

@@

\begin{frame}{Experiment(Cont'd)}
    \begin{figure}
        \centering
        \includegraphics[width=0.8\textwidth]{fig/mixed_graph_example.png}
        \caption{Example of mixed graph}
        \label{fig:mixed_graph_example}
    \end{figure}
\end{frame}

\begin{frame}{Experiment(Cont'd)}
    \begin{figure}
        \centering
        \includegraphics[width=0.8\textwidth]{fig/mixed_graph_adjacency.png}
        \caption{Adjacency matrix of previous example graph}
        \label{fig:mixed_graph_adjacency}
    \end{figure}
\end{frame}
